A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-08 main.tex
3.5 edits

A key feature which differentiates \eqref{eqn:model} from traditional linear mixed models is that in our setting, $\bZ$ is not observed directly, although we still wish to correct for it in order to remove the effects of confounding influences. A common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}\pb{; we will discuss the rationale for this approach in Section~\ref{sec:sim_env_conf}}.  This, in turn, motivates the methods by which LMM-Lasso and PC-Lasso attempt to model unobserved confounding effects using observed genetic data.

@@

It has been shown that the random effect formulation of model \eqref{eqn:model} used by LMM-Lasso can be equivalently expressed in terms of fixed effects \pb{in the absence of penalization?}. \citet{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \citet{hoffman2013correcting} used the singular value decomposition of $\bX$, which we briefly review.

Using the result of the eigendecomposition \pb{in what sense is this an eigendecomposition?  I would describe it as a Cholesky decomposition} from \eqref{eqn:k}, $\bK = p^{-1} \bR \bR^\T$ , the LMM model \eqref{eqn:lmm_reg} can be written equivalently as

@@

%------------------------------%
\subsection{Simulating environmental confounding}
\label{sec:sim_env_conf}

\pb{We need a new section title here; the next section is the one that focuses on simulation.  This one talks about a number of important concepts that go well beyond simulation.}

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis \citep{Rakitsch2012, bhatnagar2019simultaneous}.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude has the effect of ``washing out'' bias in $\bbetaHat$. To illustrate this concept, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg \bI)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. Consider our earlier example involving confounding due to air quality \pb{I feel that this should be moved up to before the mathematical derivations}. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding. Under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity.

@@

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ using the ratio $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2 \in [0, 1]$. In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates. In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure that can be inferred from $\bX$, as quantified by $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$, will impact the extent to which environmental heterogeneity confounds SNP-phenotype associations. \anna{this calculation is still having issues for the admixture and independent supbopulation data - the ratio is slightly larger for the independent data - as we would expect - but only on the order of 1e-15... correlation of Xtau with Zgamma? euclidian distance between vectors?}

@@

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered: (1) fine and (2) coarse. We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, or founding populations, which corresponds to a finer grained continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations, or founding populations, are of equal size. We note that the fine and coarse subpopulation structures are distinct from the level of admixture in the simulated data. For example, the discrete nonoverlapping subpopulations in the simulated independent subpopulation data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated admixture data. The empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ?} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding and potential estimation bias in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

@@

LMM-Lasso-ggmix's poor performance in these scenarios are likely attributable to a number of a factors such as an ill-conditioned RRM  in $n \ge p$ settings \citep{ledoit2004well}. In addition, because LMM-Lasso-ggmix attempts to estimate more parameters than LMM-Lasso-Rakitsch, its optimization problem is not necessarily convex, making it prone to getting stuck in local minima before it is able to extensively explore the coefficient path. An examination of the number of iterations required for convergence for each $\lambda$ in the coefficient path suggests that this is the case for our various simulation settings \anna{Supplementary figure ?}. Getting stuck in this manner would lead to few iterations required for each $\lambda$ value before convergence, allowing for an advantage in terms of computational speed, but poor estimation, as we see in Figure \ref{fig:eta_beta_mse}.

>>>>

A key feature which differentiates \eqref{eqn:model} from traditional linear mixed models is that in our setting, $\bZ$ is not observed directly, although we still wish to correct for it in order to remove the effects of confounding influences. A common convention is to assume $\bZ = \bX$ \citep{wang2018multiplex, lippert2011fast, yang2014advantages}; we will discuss the rationale for this approach in Section~\ref{sec:sim_env_conf}.  This, in turn, motivates the methods by which LMM-Lasso and PC-Lasso attempt to model unobserved confounding effects using observed genetic data.

@@

It has been shown that the random effect formulation of model \eqref{eqn:model} used by LMM-Lasso can be equivalently expressed in terms of fixed effects in the absence of penalization?. \citet{zhang2015principal} derived this equivalence using a probabilistic PCA formulation, while \citet{hoffman2013correcting} used the singular value decomposition of $\bX$, which we briefly review.

Using the result of the decomposition from \eqref{eqn:k}, $\bK = p^{-1} \bR \bR^\T$ , the LMM model \eqref{eqn:lmm_reg} can be written equivalently as

@@

%------------------------------%
\subsection{\anna{Decomposing (Quantifying?) the effect of environmental confounding}}
\label{sec:sim_env_conf}

In carrying out simulation studies of the LMM regression model \eqref{eqn:lmm_reg}, a common practice is to draw a new value of the quantity $\bX\bgamma$ from a $N(\boldsymbol{0}, \sigmagg \bK)$ for each replication of the analysis \citep{Rakitsch2012, bhatnagar2019simultaneous}.  We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method in the presence of environmental heterogeneity. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude has the effect of ``washing out'' bias in $\bbetaHat$.

Consider our earlier example involving confounding due to air quality. Simulating $\bgamma$ as a random quantity for each replication corresponds to subpopulation A being exposed to good air quality in one replication, and bad air quality in the next. Reporting the average across these situations is inconsistent with the question we are truly interested in, namely: what would happen if we took repeated samples from populations A and B, when these populations are subject to different environmental exposures? Those repeated samples would show a systematic bias due to the difference in air quality, but this is masked if $\bgamma$ is assumed to be random. Treating $\bgamma$ as a fixed quantity so that environmental effects are consistent across replications more closely approximates environmental confounding.

To illustrate this concept mathematically, consider the low-dimensional ordinary least squares (OLS) estimates of $\bbeta$ based on model \eqref{eqn:lmm_reg} where, without loss of generatlity, we assume $\bX_0 \bbeta_0 = \boldsymbol{0}$ for ease of notation: 
\begin{align}
    \mathbb{E}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\by) \notag \\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{E}(\bX\bbeta + \bX\bgamma + \beps) \label{eqn:expectation}\\
    &\quad \notag\\
    \mathbb{V}(\bbetaHat) &= (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\by) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bbeta + \bX\bgamma + \beps) \notag\\
    &=  (\bX^\T \bX)^{-1} \bX^\T \mathbb{V}(\bX\bgamma + \beps) \bX  (\bX^\T \bX)^{-1} \label{eqn:variance}.
\end{align}
Under the assumption that environmental effects arise from a random process, $\bgamma \sim N(\boldsymbol{0}, p^{-1} \sigmagg \bI)$, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta$, and $\mathbb{V}(\bbetaHat) = p^{-1}\sigmagg\bI + \sigmaee (\bX^\T \bX)^{-1}$, respectively. However, under the assumption that the quantity $\bgamma$ is fixed, equations \eqref{eqn:expectation} and \eqref{eqn:variance} yield $\mathbb{E}(\bbetaHat) = \bbeta + \bgamma, \mathbb{V}(\bbetaHat) = \sigmaee (\bX^\T \bX)^{-1}$, allowing us to quantify estimation bias due to environmental heterogeneity.

@@

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ \anna{and thus biases the $\bbeta$ estimates} using the ratio \anna{$||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2 \in [0, 1]$}. In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates.

In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure that can be inferred from $\bX$, as quantified by \anna{$||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2$}, will impact the extent to which environmental heterogeneity \anna{has the potential to bias} SNP-phenotype associations. \anna{We will use this metric in Section \ref{sec:results}. Given the high-dimensional nature of our data, we will approximate $(\bX^\T \bX)^{-1}$ using a low-rank approximation. We use this approach as opposed to that of a generalized inverse in order to ensure a unique solution.}

@@

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered: (1) fine and (2) coarse. We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, or founding populations, which corresponds to a finer grained continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations, or founding populations, are of equal size. We note that the fine and coarse subpopulation structures are distinct from the level of admixture in the simulated data. For example, the discrete nonoverlapping subpopulations in the simulated independent subpopulation data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated admixture data. The empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ? MAKE THESE AND MAKE THEM PRETTY} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$: Admixture ratio = 0.46412, Independent ratio = 0.98615, Vanderwas ratio = 0.64325} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding and potential estimation bias in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

@@

LMM-Lasso-ggmix's poor performance in these scenarios are likely attributable to a number of a factors such as an ill-conditioned RRM  in $n \ge p$ settings \citep{ledoit2004well}. In addition, because LMM-Lasso-ggmix attempts to estimate more parameters than LMM-Lasso-Rakitsch, its optimization problem is not necessarily convex, making it prone to getting stuck in local minima before it is able to extensively explore the coefficient path. An examination of the number of iterations required for convergence for each $\lambda$ in the coefficient path suggests that this is the case for our various simulation settings \anna{Supplementary figure ? MAKE THIS AND MAKE IT PRETTY}. Getting stuck in this manner would lead to few iterations required for each $\lambda$ value before convergence, allowing for an advantage in terms of computational speed, but poor estimation, as we see in Figure \ref{fig:eta_beta_mse}.
